---
title: "HR Analytics"
author: "Suma Krishnaprasad"
date: "July 10, 2015"
output:
  pdf_document: default
  html_document:
    highlight: tango
    theme: united
---

###Executive Summary
This population study examined employee retention in a North American Acute care Hospital over a 10 year period to determine what factors influence employee turnover. The hospital was interested in building a model that could be used to predict which employees had a higher likelihood of leaving the organization so that timely interventions could be executed to improve employee retention.

The main questions that needed to be addressed were - 

1. What factors affect voluntary employee turnover?

2. How can employee turnover be predicted?

3. How can voluntary, regrettable, avoidable employee turnover be reduced?

###HR Analytical Model
HR experts were consulted and HR theories such as Human Capital theory, Social Identity theory, Expectancy and Distributive justice theory were considered while carefully selecting the various factors that may affect employee turnover. The goal of the project was to build a simple actionable analytical model, simple enough for HR executives to understand, actionable, provide clear insights and assist in developing a retention campaign.

Employee termination was classified into the following types - 

* VTO-R-A - Voluntary turnover - Regrettable- Avoidable

* VTO-R-U - Voluntary turnover - Regrettable - Unavoidable

* VTO-N   - Voluntary turnover - NonRegrettable

* ITO     - Involuntary turnover

* RIF     - Reduction in force

The termination type of interest was VTO-R-A - Voluntary turnover - Regrettable - Avoidable

Factors selected for analysis were grouped into 3 main categories - 

**Socio-Demographic variables** - 

  * ZipCode  - Zipcode the employee resides in
  
  * Sex  - Gender of employee
  
  * Eeoc - Race/Ethnicity of employee
  
  * Age - Current age of employee
  
  * EntryAge - Age employee joined organization
  
  * CareerAge - Age at termination if terminated else current age
  
  * Education Level - 1 - No HS; 2 - Some HS; 3 - HS; 4 - GED; 5 - Some College; 6 - Associate Degree; 7 -      
                    Under Grad; 8 - Graduate; 9 - PHD; 10 - Post Doctoral
                    
  * Experience - Number of years employed prior to joining organization
  
  * Career stage - based on levinson model - 1 - <25; 2- 25-35; 3 - 35-50; 4 - 50+; 
  
  * Service Length - Factor levels based on length of total employed years - 
                    1 - <1; 2 - 1-2; 3 - 3-5; 4 - 5-10; 5 - 11-20; 6 - 20+
                                
**Work variables** - 

  * Start Rate - hourly rate at which employee started their job at organization
  
  * BaseRate - Current base hourly rate of employee
  
  * SalaryProg - Average increase in salary per employed year 
  
  * StartGrade - Grade employee started their job at
  
  * Grade - Current grade of employee
  
  * JoblevelProg - Average increased in grade per employed year
  
  * StartEval - Evaluation score of employee at start of job at organization
  
  * LastEval - Last evaluation score of employee (1 - 10 overall score)
  
  * AvgEval - Average evaluation score of employee
  
  * Tenure - Number of years employed at current organization
  
  * TotalEmployment - Total years employed since start of career
  
  * LastPromotion - Number of years since last promotion
  
  * Training - Amount spent on employee on education and training
  
  * Unscheduled - Number of unscheduled leaves taken in the last year
  
  * Shift - If the employee works on shifts (1 - No shift; 2 - Shift; 3 - Late night shifts)
  
**Psychometric variables** - 
  
  * Employee Engagement Index - Measure to determine how engaged an employee is with their work and 
                              organization. A 1 - 10 score based on employee survey and manager evaluation
  
  * Employee Commitment Index - Measure to determine how committed an employee is to the organization and 
                              their job. A 1 - 10 score based on employee survey and manager evaluation

Other factors that were important but were left out due to lack of reliable data or due to client reluctance to include in model -
  
  * Supervisor/Manager of Employee
  
  * Job fit index
  
  * Market oppurtunity index
  
  * Work/Life imbalance
  
  * Stress Index

###Exploratory Data Analysis

2000 employee records over a period of 10 years was selected for model building. The records mainly consisted of voluntary regrettable and avoidable turnover employees and several active employees randomly selected. A summary and sample of the data can be seen below -

```{r warning=FALSE,fig.width=12,fig.height=8}
library(rpart.plot)
library(rpart)
library(caret)
library(car)
library(randomForest)
setwd("~/Documents/BI-Dashboards/HRAnalytics/")
training <- read.csv("HRAnalytics_Model.csv")

dim(training)
summary(training)
head(training)
```

The main variables were grouped by category and scatter plots were constructed to analyse relationships among variables.

```{r, warning=FALSE,fig.width=12,fig.height=8}

scatterplotMatrix(~Shift+BaseRate+Grade+LastEval+LastPromotion+Age|Status,data=training)
scatterplotMatrix(~Education+StartGrade+StartRate+StartEval+Experience|Status,data=training)
scatterplotMatrix(~SalaryProg+JobLevelProg+AvgEval+Tenure+CareerAge|Status,data=training)
scatterplotMatrix(~EngagementIndex+CommitmentIndex+Unscheduled|Status,data=training)

```

From the scatterplots relationships among various variables can be seen and their impact on Status - active vs terminated employees. 

###Predictive Analysis
Since this is a binomial prediction of active vs terminated employees, a decision tree structure is the best method to construct to a prediction model.

We divide the data set into training and validation sets. The training set is then divided into training and test set. Various models are trained on the training set and tested on the test set. Then the best model is then validated on the validation set and measured for accuracy.

```{r, warning=FALSE,fig.width=12,fig.height=8}
partition <- createDataPartition(y = training$Status, p = 0.8, list = FALSE)
training <- training[partition, ]
testing <- training[-partition, ]

partition <- createDataPartition(y = training$Status, p = 0.6, list = FALSE)
traindata <- training[partition, ]
testdata <- training[-partition, ]

```


###RPart
After partitioning the data sets - the rpart method is used to construct a decision tree and check accuracy

```{r, warning=FALSE,fig.width=12,fig.height=8}
tree <- rpart(Status ~ ZipCode+Sex+Eeoc+Shift+BaseRate+Grade+Training+LastEval+StartEval+Education+StartGrade+StartRate+Experience+EngagementIndex+CommitmentIndex+Unscheduled+Tenure+Age+EntryAge+CareerAge+CareerStage+ServiceLength+SalaryProg+JobLevelProg+AvgEval+LastPromotion+TotalEmployment, data=traindata)
tree
printcp(tree)
prp(tree)
```

As can be seen with this method there is a 45% error rate, which is not a good model for our prediction.

###RandomForest

Next the randomforest method is applied. Randomforest is one of the best classification tree predictive models in the industry today.

```{r, warning=FALSE}
model.rf<-randomForest(traindata[,3:29],traindata$Status)
model.rf

training_pred <- predict(model, traindata)
confusionMatrix(training_pred, traindata$Status)

testdata_pred <- predict(model.rf, testdata)
confusionMatrix(testdata_pred, testdata$Status)

varImpPlot(model.rf,sort=TRUE,n.var=27)
```

As can be seen - the accuracy of the training data prediction is about 91% with an error rate of 18% and the test data accuracy is 80%. This is better than the previous result, but still not quite up to par for our requirements.

###RandomForest with Bootstrapping
Next RandomForest with a 10-fold bootstrapping technique with replace is applied.

```{r, warning=FALSE}
model <- train(traindata[,3:29],traindata$Status, method = "rf", prox = TRUE, trControl = trainControl(method = "cv", number = 10, allowParallel = TRUE))
training_pred <- predict(model, traindata)
confusionMatrix(training_pred, traindata$Status)

testdata_pred <- predict(model, testdata)
confusionMatrix(testdata_pred, testdata$Status)

varImp(model)
```

Finally we will apply the model against our validation set to test our prediction -
Confusion matrix along with a list of employees that were incorrectly predicted is included for further analysis.

```{r, warning=FALSE}
testing_pred <- predict(model, testing)
confusionMatrix(testing_pred, testing$Status)

testing.pred<-cbind(testing,testing_pred)
subset(testing.pred, Status!=testing_pred, select=c("EmployeePPPER","Status","testing_pred"))
```

We obtain a **92% accuracy** on the validation dataset which is a pretty good prediction rate.

As can be seen from both randomforest models the top 5 factors affecting terminations are - 

* Baserate

* Age

* LastPromotion

* Zipcode

* Tenure

These 5 variables have a much lar

Lets plot these 5 variables against status and check their relationships again

```{r, warning=FALSE,fig.width=12,fig.height=8}
scatterplotMatrix(~Status+BaseRate+Age+LastPromotion+ZipCode+Tenure, data=testing)
```

###Conclusion
From the above matrix it can be concluded that **younger employees (40 years or below)** with **lower BaseRates (<$20/hr)**, who have **not received a promotion in a while (last 5 years)**, who **commute long distances to work** and have **worked for the organization 5 years or less** tend to leave the organization

